Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is a merged and final version of two conference papers, \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}, with a unified and streamlined presentation and expanded background materials.}

### 2020-08-22 journal paper/code_abstract.tex
add CODE abstract

@@

\subsection{Our model: competing bandits}
\label{sec:intro-model}

@@

While we build on a vast and active research area, we present sufficient background in Appendix~\ref{app:bg} so as to keep this paper accessible to non-specialists.

\subsection{Our Results}
\label{sec:intro-results}

@@

All results extend to a much more general version of the multi-armed bandit problem in which the principal may observe additional feedback before and/or after each decision, as long as the feedback distribution does not change over time. In most results, principal's utility may depend on both the market share and agents' rewards.

\begin{figure}
\begin{center}
\begin{tikzpicture}
      \draw[->] (-.5,0) -- (3,0) node[right] {Competitiveness};
      \draw[->] (0,-.5) -- (0,3) node[above] {Better algorithms};
      \draw[scale=0.6,domain=0.5:4.5,smooth,variable=\x,blue, line width=0.3mm] plot ({\x},{4.5 - (\x - 2.5)^2});
      % \draw[scale=0.5,domain=-3:3,smooth,variable=\y,red]  plot ({\y*\y},{\y});
 \end{tikzpicture}

\caption{Inverted-U relationship between competitiveness and algorithms.}
\label{fig:inverted-U}
\end{center}
\end{figure}

\xhdr{Economic interpretation: the inverted-U relationship.}
Interpreting the adoption of better algorithms as ``innovation", our findings can be framed in terms of the inverted-U relationship between competition and innovation, as in Figure~\ref{fig:inverted-U}.%
\footnote{Here, ``innovation" refers to adoption of a better technology at a substantial R\&D expense to a given firm. It is not salient whether similar ideas and/or technologies already exist elsewhere. Adoption of exploration algorithms tends to require substantial R\&D effort in practice, even if the algorithms themselves are well-known in the research literature \citep[\eg see][]{DS-arxiv}.}
This is a well-established concept in the economics literature, dating back to \cite{Schumpeter-42}, whereby too little or too much competition is bad for innovation, but intermediate levels of competition tend to be better \citep[\eg][]{aghion2005competition,Vives-08}.

Our decision rules differ in terms of rationality: from fully rational decisions with \HardMax to relaxed rationality with \HardMaxRandom to an even more relaxed rationality with \SoftMaxRandom. The same distinctions also control the severity of competition between the principals: from cut-throat competition with \HardMax to a more relaxed competition with \HardMaxRandom, to an even more relaxed competition with \SoftMaxRandom. Indeed, with \HardMax you lose all customers as soon as you fall behind in performance, with \HardMaxRandom you get some small market share no matter what, and with \SoftMaxRandom you are further guaranteed a market share close to $\tfrac12$ as long as your performance is not much worse than the competition. The uniform choice among principals corresponds to no rationality and no competition. While agents' rationality and severity of competition are often modeled separately in the literature, it is not unusual to have them modeled with the same ``knob" \cite[\eg][]{Gabaix-16}.

We identify the inverted-U relationship driven by the rationality/competitiveness distinctions outlined above: from \HardMax to \HardMaxRandom to \SoftMaxRandom to \Uniform. We also find another, technically different, inverted-U relationship which zeroes in on the \HardMaxRandom model. We vary rationality/competitiveness inside this model, and track the marginal utility of switching to a better algorithm.

These inverted-U relationships are driven by different aspects in our model than the ones in the existing literature in economics. The latter focuses on the tradeoff between the R\&D costs and the benefits that the improved technology provides in the competition. In our case, the barriers for innovations arise entirely from the reputational consequences of exploration in competition, even in the absence of R\&D costs.

@@

\item We investigate the algorithms' performance ``in isolation", \ie without competition. We suggest a new performance measure to explain why the greedy algorithm is sometimes not the best strategy under high levels of competition.\footnote{Note the discrepancy with our theoretical results on \HardMax: there, the greedy algorithm is always the best strategy, mainly because it is aware of the Bayesian prior (whereas in the simulations the prior is not available).}
     We find that mean reputation -- arguably, the most natural performance measure ``in isolation" -- is sometimes \emph{not} a good predictor for the outcomes under competition.

@@

\ascomment{One downside of the story above: we may need to explain why we didn't simulate \SoftMaxRandom.}

\ascomment{The xhdr below is basically from Guy, with some rewordings by Alex.}

\xhdr{Economic interpretation: network effects of data.}
Our model speaks to policy discussions on regulating data-intensive digital platforms \citep{furman2019unlocking, scott2019committee}, and particularly to the ongoing debate on the role of data in the digital economy. One fundamental question in this debate is whether data can serve a similar role as traditional ``network effects", creating scenarios when only one firm can function in the market \citep{Rysman09, jullien2019economics}.
%whereby, when these effects are present, in many cases only one firm can function in the market, leading to competition \emph{for} the market being more important than competition \emph{in} the market \citep{Rysman09, jullien2019economics}.
The death spiral/amplification effects mentioned above have a similar flavor as network effects: a relatively small amount of exploration (resp., data advantage)  gets amplified under competition and causes the firm to be starved of users (resp., take over most of the market).
%\gaedit{We further find that a small data advantage for one firm gets amplified under competition and leads to that firm taking the entire market, showing that data can provide a similar incumbency advantage as those provided by traditional network effects and can serve as a barrier to entry in online markets.}
A distinctive feature of our approach is that we explicitly model the learning problem of the firms and consider them deploying algorithms for solving this problem.  Thus, we do not explicitly model the network effects, but they arise endogenously from our setup.

Our results highlight that understanding the performance of learning algorithms in isolation does not necessarily translate to understanding their impact in competition, precisely due to the fact that competition leads to the endogenous generation of data observed by the firms. Approaches such as \citet{lambrecht2015can, bajari2018impact, varian2018artificial} argue that the diminishing returns to scale and scope of data in isolation mitigate such data feedback loops,
%as non-existent
but ignore the differences induced by learning in isolation versus under competition. Furthermore, explicitly incorporating the interaction between learning technology and data creation allows us to speak on how data advantages are characterized and amplified not only by data quantity, but also the increased data quality gathered by better learning algorithms.

\ascomment{"Econ interpretation" xhdrs fit incely in "our results", as opposed to a separate subsection. Because: we can explain the "inverted-U" stuff right after "theory", before we put the simulation results on the stack. And the "network effects" stuff goes nicely right after the numerical results, without a big subsection break.}

\subsection{Further discussion}
\label{sec:discussion}

\xhdr{Significance.}
Our theory takes a basic Bayesian approach, a standard perspective in economic theory, and discovers several strong asymptotic results. The numerical simulations provide a more nuanced and ``non-asymptotic" perspective. In essence, we look for substantial effects within relevant time scales. In fact, we start our investigation by determining what time scales are relevant in the context of our model.

\ascomment{The next para makes a point that seems important (to me). Is this point clear?}

Our study has a dual purpose: (i) shed light on real-world implications of some typical scenarios, and (ii) investigate the space of models for describing the real world. As an example to clarify the latter point, consider the \HardMax model with simultaneous entry. It is is not necessarily the most realistic model, but arguably the most natural one to study \emph{a priori}. However, our results elucidate the need for more refined models that allow for ``free exploration" (\eg via random agents or early entry).

\xhdr{Modeling.}
Our models are stylized in several important  respects. Firms compete only on the quality of service, rather than, say, pricing or the range of products. Agents are myopic: they do not worry about how their actions impact their future utility.\footnote{In particular, agents do not attempt to learn over time, to second-guess or game future agents, or to manipulate the principals' learning algorithms. This is arguably typical in practice, in part because one agent's influence tends to be relatively small.}  Various performance signals available to the users, from personal experience to word-of-mouth to consumer reports, are abstracted as persistent ``reputation scores", and further simplified to average performance over a time window. On the machine learning side, our setup is geared to distinguish between ``simple" vs. ``better" vs. ``smart" bandit algorithms; we are not interested in state-of-art algorithms for very realistic bandit settings.

We consider two extremes: a simple Bayesian model with full Bayesian rationality and no performance signals, and a simple frequentist model with reputation scores and no prior knowledge or beliefs. For the theoretical results, the ``no-performance-signals" assumption makes agents' behavior independent of a particular realization of the prior. Therefore, we summarize each learning algorithm via its Bayesian-expected rewards, not worrying about its detailed performance on particular realizations of the prior. Such summarization is essential for formulating lucid and general analytic results, let alone proving them. It is unclear how to incorporate performance signals in a theoretically tractable model. For the numerical results, the \ExptsModel accounts for competition in a more direct way, allows to separate reputation vs. data advantage, and makes our model amenable to numerical simulations. Indeed, simulating the intricate interplay of learning dynamics and Bayesian priors appears computationally intractable.

\asedit{Principals' utilities (and, accordingly, bandit rewards) are not discounted with time.%
\footnote{However, our main theoretical result on the greedy algorithm extends to time-discounted rewards, see Section~\ref{sec:theory-extensions}.}
 While much of the early work focuses on  time-discounted bandits \citep{Gittins-book11}, non-discounted  formulations are prevalent in the bandits literature over the past two decades \citep{slivkins-MABbook,LS19bandit-book}, and better correspond to practical deployments \citep[\eg][]{DS-arxiv}. Moreover, the distinctions between better and worse bandit algorithms are not as well-understood under time-discounting.}

\xhdr{Challenges.}
Much of the challenge in the theoretical results, both conceptual and technical, was in setting up the model and the theorems. Apart from zeroing in on the \TheoryModel, it was crucial to interpret the results and intuitions from the literature on multi-armed bandits so as to formulate meaningful and productive assumptions on bandit algorithms and Bayesian priors.

The numerical investigation is quite challenging even with a stylized model such as ours. An ``atomic experiment" is a competition game between a given pair of bandit algorithms, in a given competition model, on a given instance of a multi-armed bandit problem (and each such experiment is run many times to reduce variance).
Accordingly, we have a three-dimensional space of atomic experiments one needs to run and interpret: \{pairs of algorithms\} x \{competition models\} x \{bandit instances\}, and we are looking for findings that are consistent across this entire space. It is essential to keep each of the three dimensions small yet representative. In particular, we need to capture a huge variety of bandit instances with only a few representative examples. Further, we need a succinct and informative summarization of results within one atomic experiment and across multiple experiments (\eg see Table~\ref{sim_table}).

While amenable to simulations, the \ExptsModel appears difficult to analyze. This is for several reasons:
%
(i) intricate feedback loop from performance to reputations to users to performance;
%
(ii) mean reputation, most connected to our intuition, is sometimes a bad predictor in competition (see Sections~\ref{sec:isolation} and~\ref{sec:revisited}); and
%
(iii)
mathematical tools from regret-minimization would only produce ``asymptotic" results, which do not seem to suffice. Given the theoretical results on the \TheoryModel, and the fact that we are in the realm of stylized economic models, we believe that resolving first-order theoretical questions about the \ExptsModel would not add much value to this paper.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

>>>>

\documentclass[10pt]{article}

\usepackage{color,
setspace,sectsty,comment,footmisc,caption,
pdflscape,subcaption,array,hyperref,adjustbox,threeparttable
}

\usepackage{fullpage}
%\usepackage{geometry}
%\geometry{left=1.0in,right=1.0in,top=1.0in,bottom=1.0in}

\usepackage{subfiles}
\usepackage{libertine}
\usepackage[title,titletoc]{appendix}
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage[round]{natbib} % cannot use with AAAI! A.S.

\usepackage{float}
\usepackage{sgame, tikz} % Game theory packages
\usepackage{algorithm,algpseudocode}
\usepackage{makecell}
 \usepackage{multirow}
 \usepackage{graphicx}

\usepackage{booktabs} % For formal tables

%\theoremstyle{definition}
%\newtheorem{definition}{Definition}
\newtheorem{finding}{Finding}
%\newtheorem{conjecture}{Conjecture}
\newtheorem{puzzle}{Puzzle}
\captionsetup{font=footnotesize}

\usepackage{slivkins-setup,slivkins-theorems}

% model variants
\newcommand{\TheoryModel}{Bayesian-choice model\xspace}
\newcommand{\ExptsModel}{reputation-choice model\xspace}

% Commands from EC 19 paper
\newcommand{\term}[1]{\ensuremath{\mathtt{#1}}\xspace}

% shorthand for algorithms
\newcommand{\TS}{\term{TS}}    % Thompson Sampling
\newcommand{\DEG}{\term{BEG}}  % "Dynamic eps-greedy"
\newcommand{\DG}{\term{BG}}    % "Dynamic Greedy"

% shorthand for models
\newcommand{\HMR}{\term{HMR}} % HardMaxRandom
\newcommand{\HM}{\term{HM}}    % HardMax

\newcommand{\Beta}{\term{Beta}} % for Beta distribution
\newcommand{\Eeog}{\term{EoG}} % shorthand for "effective end of game"

\newcommand{\MRV}{mean reward vector\xspace} % mean reward vector
\newcommand{\MRVs}{mean reward vectors\xspace} % mean reward vector

% commands from "theory"

% in connection with "a little greedy goes a long way"
%     stuff for the "greedy modification".
\newcommand{\gr}{\texttt{gr}}
\newcommand{\BIRgr}{\BIR^{\gr}}
\newcommand{\alggr}{\alg[\gr]}
\newcommand{\rewgr}{\rew_{\gr}}

\DeclareMathOperator*{\Expectation}{\mathbb{E}}
\newcommand{\Ex}[2]{\Expectation_{#1}\left[#2\right]}
\newcommand{\FiniteGame}{finite competition game\xspace}

% notation
%%% advanced notation
\newcommand{\OPT}{\term{OPT}}
\newcommand{\rew}{\term{rew}}  % Bayesian-expected reward after n local rounds
\newcommand{\PMR}{\term{PMR}} % posterior mean reward
\newcommand{\support}{\term{support}}

\newcommand{\BIR}{\term{BIR}} % Bayesian Instantaneous Regret

% response function
\newcommand{\respF}{f_{\term{resp}}}
\newcommand{\respEps}{\eps_\term{resp}}

\newcommand{\BReg}{\term{BReg}}
\newcommand{\HardMax}{\term{HardMax}}
\newcommand{\HardMaxRandom}{\term{HardMax\&Random}}
\newcommand{\SoftMaxRandom}{\term{SoftMax}}
\newcommand{\Uniform}{\term{Uniform}}
\newcommand{\Random}{\term{Random}}

% algorithms
\newcommand{\StaticGreedy}{\term{StaticGreedy}}
\newcommand{\DynGreedy}{\term{BayesianGreedy}}
\newcommand{\DynamicGreedy}{\DynGreedy}
\newcommand{\DynamicEpsGreedy}{\term{BayesianEpsilonGreedy}}
\newcommand{\Thompson}{\term{ThompsonSampling}}

% "monotone algorithms"
\newcommand{\bmonotone}{Bayesian-monotone\xspace}
\newcommand{\bmonotonicity}{Bayesian-monotonicity\xspace}

\newcommand{\termSub}[2]{\ensuremath{\mathtt{#1}_{#2}}\xspace}
\newcommand{\alg}[1][]{\termSub{alg}{#1}}
%\newcommand{\prin}[1][]{\termSub{prin}{#1}}  % principal
\newcommand{\agent}[1][]{\termSub{agent}{#1}}

% priors and posteriors
\newcommand{\prior}{\ensuremath{\mP}\xspace}
\newcommand{\priorMu}{\ensuremath{\prior_\mathtt{mean}}\xspace}
\newcommand{\posteriorN}[2]{\mN_{#1,#2}}  % \posteriorN{principal}{round}

% rationality / innovation / competition
\newcommand{\termTXT}[1]{{\em {#1}}\xspace}

\newcommand{\rationality}{\termTXT{rationality}}
\newcommand{\Rationality}{\termTXT{Rationality}}
\newcommand{\innovation}{\termTXT{innovation}}
\newcommand{\Innovation}{\termTXT{Innovation}}
\newcommand{\competition}{\termTXT{competition}}
\newcommand{\Competition}{\termTXT{Competition}}
\newcommand{\competitiveness}{\termTXT{competitiveness}}
\newcommand{\exploration}{\termTXT{exploration}}
\newcommand{\Exploration}{\termTXT{Exploration}}

% a very useful package for edits and comments, from David Kempe (USC)
\usepackage[suppress]{color-edits}
%\usepackage[suppress]{color-edits}  % use this to suppress the package
\addauthor{as}{red}    % as for Alex
\addauthor{ga}{blue}  % ga for Guy
\addauthor{sw}{orange} % sw for Steven
\addauthor{ym}{madenta} % sw for Steven
% e.g. for Alex, provides \asedit{}, \ascomment{} and \asdelete{}.

\setlength{\tabcolsep}{8pt} % Default value: 6pt
\renewcommand{\arraystretch}{1.5} % Default value: 1

\begin{document}

%\begin{titlepage}

\title{Competing Bandits:\\
The Perils of Exploration under Competition%
\thanks{This is a merged and final version of two conference papers,
\citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19},
with a unified and streamlined presentation and expanded background materials.}}
% \textit{Competing Bandits: Learning under Competition}, at Innovations in Theoretical Computer Science 2018, and \textit{The Perils of Exploration under Competition: A Computational Modeling Approach}, at ACM Economics and Computation 2019.}}

\author{Guy Aridor%
\footnote{Columbia University, Department of Economics. Email: g.aridor@columbia.edu}
\and
%\rule{0.2in}{0pt}
Yishay Mansour%
\footnote{Google and Tel Aviv University, Department of Computer Science. Email: mansour.yishay@gmail.com}
\and
%\rule{0.2in}{0pt}
Aleksandrs Slivkins%
\footnote{Microsoft Research New York City. Email: slivkins@microsoft.com}
\and
%\rule{0.2in}{0pt}
Zhiwei Steven Wu%
\footnote{University of Minnesota - Twin Cities, Department of Computer Science. Email: zsw@umn.edu\newline
Part of the research was done when Z.S. Wu was an intern and a postdoc at Microsoft Research NYC.
}}
\date{August 2020}
\maketitle

@@

\subsection*{Our model: competing bandits}
\label{sec:intro-model}

@@

\subsection*{Our Results}
\label{sec:intro-results}

@@

@@

@@

\vspace{0.2in}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below.
%

\addcontentsline{toc}{section}{References}

\bibliographystyle{plainnat}
%\begin{small}
\bibliography{bib-abbrv,bib-ML,refs,bib-bandits,bib-AGT,bib-slivkins, bib-random}
%\end{small}

\clearpage
%\subfile{content/old-intro}

\end{document}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
